\documentclass{article}
\usepackage{tokcycle}[2021-03-10]
\usepackage{xcolor}
\newcounter{wordcount}
\newcounter{lettercount}
\newcounter{wordlimit}
\newif\ifinword
% USER PARAMETERS
\newif\ifrunningcount
\newif\ifsummarycount
\def\limitcolor{red}
\setcounter{wordlimit}{0}
%%
\makeatletter
% \tc@defx is like \def, but expands the replacement text once prior to assignment
\newcommand\addtomacro[2]{\tc@defx#1{#1#2}}
\newcommand\changecolor[1]{\tctestifx{.#1}{}{\addcytoks{\color{#1}{}}%
  \tc@defx\currentcolor{#1}}}
\makeatother
\newcommand\dumpword{%
  \addcytoks[1]{\accumword}%
  \ifinword\stepcounter{wordcount}
    \ifrunningcount\addcytoks[x]{$^{\thewordcount,\thelettercount}$}\fi
    \ifnum\thewordcount=\value{wordlimit}\relax\changecolor{\limitcolor}\fi
  \fi%
  \inwordfalse
  \def\accumword{}}
\newcommand\addletter[1]{%
  \tctestifcatnx A#1{\stepcounter{lettercount}\inwordtrue}{\dumpword}%
  \addtomacro\accumword{#1}}
\xtokcycleenvironment\countem
  {\addletter{##1}}
  {\dumpword\groupedcytoks{\processtoks{##1}\dumpword\expandafter}\expandafter
    \changecolor\expandafter{\currentcolor}}
  {\dumpword\addcytoks{##1}}
  {\dumpword\addcytoks{##1}}
  {\stripgroupingtrue\def\accumword{}\def\currentcolor{.}
    \setcounter{wordcount}{0}\setcounter{lettercount}{0}}
  {\dumpword\ifsummarycount\tcafterenv{%
    \par(Wordcount=\thewordcount, Lettercount=\thelettercount)}\fi}
\begin{document}
\countem 
This is a test if word counting occurs. \footnote{Whether it occurs in footnotes.}

\section{And whether it occurs in headings.}

It sort of works but it does not include citations, which will be hard to do before they are compiled. I will probably ultimately need to use Perl and TexCount.


\endcountem
I know there were \thewordcount{} words and \thelettercount{} letters in the
  prior sentence.

\bigskip\setcounter{wordlimit}{7}
\countem 
This is a test if color changes after seven words.
\endcountem

\bigskip\runningcounttrue
\countem
This is a running count.

But...punctuation does not count as characters.
\endcountem

\bigskip
\countem 
This \textbf{is a} running count. |\textsc{Skipping macros 
\rule{1ex}{1.5ex}/other text}|

But...\textit{punctuation does} not count as characters.
\endcountem

\bigskip\runningcountfalse\summarycounttrue\setcounter{wordlimit}{125}
\countem
As any dedicated reader can clearly see, the Ideal of 
practical reason is a representation of, as far as I know, the things
in themselves; as I have shown elsewhere, the phenomena should only be
used as a canon for our understanding. The paralogisms of practical
reason are what first give rise to the architectonic of practical
reason. As will easily be shown in the next section, reason would
thereby be made to contradict, in view of these considerations, the
Ideal of practical reason, yet the manifold depends on the phenomena.
Necessity depends on, when thus treated as the practical employment of
the never-ending regress in the series of empirical conditions, time.
Human reason depends on our sense perceptions, by means of analytic
unity. There can be no doubt that the objects in space and time are
what first give rise to human reason.

Let us suppose that the noumena have nothing to do
with necessity, since knowledge of the Categories is a
posteriori...
\endcountem
\end{document}